---
title: "Exploratory Data Analysis of Red Wine"
author: "Marco Rodriguez"
date: "December 2, 2018"
output: html_document
---


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

knitr::opts_chunk$set(fig.width=9,fig.height=5,fig.path='Figs/',
                      fig.align='center',tidy=TRUE,
                      echo=FALSE,warning=FALSE,message=FALSE)

# load pachages
library(ggplot2) # Need this for plotting
library(gridExtra) # Need this for grid.arrange
library(dplyr) # To group data
library(scales) # For log10_trans
library(memisc) # For the identity function in modelling
library (knitr) # To output rmarkdown to html
library (rmarkdown)
library(GGally) # ggpairs
library(psych)
theme_set(theme_bw())

```

```{r echo=FALSE, Load_the_Data}
# Load the Data
rwine <- read.csv('wineQualityReds.csv', sep = ',')

```

```{r echo=FALSE, Dataset_First_Rows}
# data set first rows
head(rwine)
```

```{r echo=FALSE, Dataset_Summary}
# stats summary
summary(rwine)
```

```{r echo=FALSE, Dataset_Dim}
# Data set dimensions
dim(rwine)
```

```{r echo=FALSE, Dataset_Col_Names}
# Data set columns names
names(rwine)
```

# Univariate Plots Section

## Histograms to show the distribution of the variable values:

```{r echo=FALSE, Univariate_Hist_FixedAcidity}
# histogram for fixed.acidity
qplot(fixed.acidity, data = rwine, fill = I('#008B8B'), binwidth = .3)

```

```{r echo=FALSE, Univariate_Summary_FixedAcidity}
# stats summary for fixed.acidity
summary(rwine$fixed.acidity)

```

> Max values for fixed acidity is 15.90 and min is 4.60. There are some outliers

```{r echo=FALSE, Univariate_Hist_VolatileAcidity}
# histogram for volatile.acidity
qplot(volatile.acidity, data = rwine, fill = I('#008B8B'), binwidth = .01)

```

```{r echo=FALSE, Univariate_Summary_Volatile}
# stats for volatile.acidity
summary(rwine$volatile.acidity)

```

> Max value is 1.58 and min 0.12. There are some outliers.

```{r echo=FALSE, Univariate_Hist_CitricAcid}
# histogram for citric.acid
qplot(citric.acid, data = rwine, fill = I('#008B8B'), binwidth = .01)

```

```{r echo=FALSE, Univariate_Summary_CitricAcid}
# stats for citric.acid
summary(rwine$citric.acid)

```

> The min value for citric.acid is 0.000 and max value is 1.00. Histogram illustrate that most those values are equal to 0.000. Some outliers too.

```{r echo=FALSE, Univariate_Hist_ResidualSugar}
# histogram for residual.sugar
qplot(residual.sugar, data = rwine, fill = I('#008B8B'), binwidth = .3)

```


```{r echo=FALSE, Univariate_Summary_residual.sugar}
# stats for residual.sugar
summary(rwine$residual.sugar)

```

> We notice an irregular shape for residual.sugar. Max value is 15.50 and min 0.900. There are some outliers.

```{r echo=FALSE, Univariate_Hist_Chlorides}
# histogram for chlorides
qplot(chlorides, data = rwine, fill = I('#008B8B'), binwidth = .01)

```

```{r echo=FALSE, Univariate_Summary_Chlorides}
# stats for chlorides
summary(rwine$chlorides)

```

> The min value is 0.012 and max value is 0.6. Histogram shows that most value count are found 0.04 and 1.4. There are some outliers.

```{r echo=FALSE, Univariate_Hist_free.sulfur.dioxide}
# histogram for free.sulfur.dioxide
qplot(free.sulfur.dioxide, data = rwine, fill = I('#008B8B'), binwidth = 1)

```


```{r echo=FALSE, Univariate_Summary_free.sulfur.dioxide}
# stats for free.sulfur.dioxide
summary(rwine$free.sulfur.dioxide)

```

> There are some outliers. Most value count are found within 2 and 20

```{r echo=FALSE, Univariate_Hist_total.sulfur.dioxide}
# histogram for total.sulfur.dioxide
qplot(total.sulfur.dioxide, data = rwine, fill = I('#008B8B'), binwidth = 3)

```


```{r echo=FALSE, Univariate_Summary_total.sulfur.dioxide}
# stats for total.sulfur.dioxide
summary(rwine$total.sulfur.dioxide)

```

> Plot seens right skewed. There are some outliers. Max value is 289 and min 6.00

```{r echo=FALSE, Univariate_Hist_density}
# histogram for density
qplot(density, data = rwine, fill = I('#008B8B'), binwidth = .001)

```


```{r echo=FALSE, Univariate_Summary_density}
# stats for density
summary(rwine$density)

```

> Plot seens normal distribute. Max value is 1.0037 and min 0.9901

```{r echo=FALSE, Univariate_Hist_pH}
# histogram for volatile.acidity
qplot(pH, data = rwine, fill = I('#008B8B'), binwidth = .03)

```


```{r echo=FALSE, Univariate_Summary_pH}
# stats for sulpahtes
summary(rwine$pH)

```

> Plot seems noral distribute with some outliers.

```{r echo=FALSE, Univariate_Hist_sulphates}
# histogram for sulphates
qplot(sulphates, data = rwine, fill = I('#008B8B'), binwidth = .03)

```


```{r echo=FALSE, Univariate_Summary_sulphates}
# stats for sulphates
summary(rwine$sulphates)

```

> There are some outliers. Most value count fall within 0.3 and 0.9.

```{r echo=FALSE, Univariate_Hist_alcohol}
# histogram for sulphates
qplot(alcohol, data = rwine, fill = I('#008B8B'), binwidth = .3)

```

```{r echo=FALSE, Univariate_Summary_alcohol}
# stats for alcohol
summary(rwine$alcohol)

```

> Plot seems right skewed. There are some outliers.

```{r echo=FALSE, Univariate_Hist_quality}
# histogram for quality
qplot(quality, data = rwine, fill = I('#008B8B'), binwidth = .1)

```

```{r echo=FALSE, Univariate_Summary_quality}
# stats for quality
summary(rwine$quality)

```

> Quality has a small proportion in the categories 3, 4, 8, 9 and the highest proportion in the categories 5, 6, and 7. None in the categories 1, 2, and 10.

## Observations regarding variables: 

> Residual sugar has a positive skewed distribution.

> Alcohol does not have noticible outliers and it has an irregular distribution.

> Fixed acidity, volatile acidity, free sulphur dioxide, density and citric acid have outliers. 


```{r echo=FALSE, New_Variable}
# for easy understanding I would like to add a new variable to the dataset.
# This new variable will categorize the red wine quality in inferior average superior

rwine$quality.factor <- factor(rwine$quality)
rwine$qc.cat <- NA
rwine$qc.cat <- ifelse(rwine$quality >= 7, 'superior', NA)
rwine$qc.cat <- ifelse(rwine$quality <= 4, 'inferior', rwine$qc.cat)
rwine$qc.cat <- ifelse(rwine$quality == 5, 'average', rwine$qc.cat)
rwine$qc.cat <- ifelse(rwine$quality == 6, 'average', rwine$qc.cat)

rwine$qc.cat <- factor(rwine$qc.cat, levels = c("inferior", "average", "superior"))

```

# Univariate Analysis

### What is the structure of your dataset?

> 1599 observation and 13 variables or characteristics.

### What is/are the main feature(s) of interest in your dataset?

> Quality is one of my main characteristic. The quality of a wine is closely linked to its properties such as alcohol, density, pH.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

> Other features in the data set that will help support the investigation are are alcohol content, pH and density. 

### Did you create any new variables from existing variables in the dataset?

>  Yes, a new categorical variable called qc.cat to classify the wines into categories according to the quality value for each register.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

> Alcohol, volatile.acidity, citric.acid, residual.sugar, ph has an irregular shaped distribution. No, I didn't.

# Bivariate Plots Section

> After doing some research on the internet. Red wine quality is affect by his alcohol level, acidity, density, and pH.

## Plots for the main feature(s) of interest (Alcohol, sulphates, density, citric.acid, fixed.acidity)

```{r echo=FALSE, Bivariate_Plots}

# qc.cat and alcohol
qc.cat_alcohol <- qplot(x = qc.cat, y = alcohol, 
      data = rwine,
      geom = "boxplot")

# qc.cat and sulphates
qc.cat_sulphates <- qplot(x = qc.cat, y = sulphates, 
      data = rwine,
      geom = "boxplot")

# qc.cat and density
qc.cat_density <- qplot(x = qc.cat, y = density, 
      data = rwine,
      geom = "boxplot")

# qc.cat and citric.acid
qc.cat_citric <- qplot(x = qc.cat, y = citric.acid, 
      data = rwine,
      geom = "boxplot")

# qc.cat and fixed.acidity
qc.cat_fixed <- qplot(x = qc.cat, y = fixed.acidity, 
      data = rwine,
      geom = "boxplot")

# display plots in one fig
grid.arrange(qc.cat_alcohol, qc.cat_sulphates, qc.cat_density, qc.cat_citric, qc.cat_fixed, ncol = 3)
```

> Quality increases with high levels of alcohol, fixed.acidity, density, sulphates, citric.acid.

```{r echo=FALSE, Bivariate_Sctter_Plots_QC_Alcohol}
# Scatterplot Review of quality and alcohol
ggplot(aes(x = quality, y = alcohol), data = rwine) +
  geom_jitter(alpha = 1/3) +
  ggtitle('Scatterplot Review of quality and alcohol')
```

> red wines within cat 5 and cat 6 are the one with highest alcohol content. 

```{r echo=FALSE, Bivariate_Scatter_Plots_QC_DEnsity}
# Scatterplot Review of quality and alcohol
ggplot(aes(x = quality, y = density), data = rwine) +
  geom_jitter(alpha = 1/3) +
  ggtitle('Scatterplot Review of quality and density')
```

> In this plot we observe that those wine that fall in the categorie 5 and 6 have the highest density. It is low in cat 3, cat4, cat7, and cat8

```{r echo=FALSE, Bivariate_Sctter_Plots_QC_citric.acid}
# Scatterplot Review of quality and citric.acid
ggplot(aes(x = quality, y = citric.acid), data = rwine) +
  geom_jitter(alpha = 1/3) +
  ggtitle('Scatterplot Review of quality and citric.acid')

```

> We find citric acid level are the low in cat 3, cat 4, and cat8. Cat 5 and cat 6 have the highest levels.


```{r echo=FALSE, Bivariate_Scatter_Plots_QC_sulphates}
# Scatterplot Review of quality and sulphates
ggplot(aes(x = quality, y = sulphates), data = rwine) +
  geom_jitter(alpha = 1/3) +
  ggtitle('Scatterplot Review of quality and sulphates')

```

>  We find sulphates level are the low in cat 3, cat 4, and cat8. Cat 5 and cat 6 have the highest levels.

```{r echo=FALSE, warning=FALSE,message=FALSE, Multivariate_Plots_1}
# Scatterplot Review of quality and sulphates
sc_sulphates_qc <- ggplot(aes(x = quality, y = sulphates), data = rwine) +
  geom_jitter(alpha = 1/5) +
  xlim(3,quantile(rwine$quality, 0.99))+ 
  ylim(0,quantile(rwine$sulphates, 0.99))+
  stat_smooth(method = 'lm') + 
  ggtitle('Scatterplot Review of quality and sulphates')

# display plot
sc_sulphates_qc

```

> sulphates increases the quality of wine. 

> Plots below shows that consumers prefer wine with high level of alcohol, acidity, and density. 

## Plots for not the main feature(s) of interest (pH, residual.sugar, chlorides, free.sulfur.dioxide, total.sulfur.dioxide)

```{r echo=FALSE, Pairwise_scatterplots_0}

# quality and ph
sc_ph <- ggplot(aes(x = quality, y = pH), 
        data = rwine) +
  geom_jitter( alpha = .5)  +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)

# quality and residual.sugar
sc_residual <- ggplot(aes(x = quality, y = residual.sugar), 
        data = rwine) +
  geom_jitter( alpha = .5)  +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)

# quality and chlorides
sc_chlorides <- ggplot(aes(x = quality, y = chlorides), 
        data = rwine) +
  geom_jitter( alpha = .5)  +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)

# quality and free.sulfur.dioxide
sc_fsulfur <- ggplot(aes(x = quality, y = free.sulfur.dioxide), 
        data = rwine) +
  geom_jitter( alpha = .5)  +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)

# quality and toatl.sulfur.dioxide
sc_tsulfur <- ggplot(aes(x = quality, y = total.sulfur.dioxide), 
        data = rwine) +
  geom_jitter( alpha = .5)  +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)

# display plots in fig
grid.arrange(sc_ph, sc_residual, sc_chlorides, sc_fsulfur, sc_tsulfur, ncol = 3)

```

> As we can see in the plots above these features have some correlation with important wine features. It is noticible that cat 5 and cat6 have the concentration of those variables.

### Correlation between Chemical Attributes

```{r echo=FALSE, warning=FALSE, message=FALSE, Scatterplots_ph_fixed.acidity_Cor}
# Scatterplots ph and fixed.acidity Correlation
fixed_pH <- ggplot(aes(pH, fixed.acidity), data = rwine)+ geom_jitter()+
  xlim(3,quantile(rwine$pH, 0.99))+ 
  ylim(6,quantile(rwine$fixed.acidity, 0.99))+
  stat_smooth(method='lm')

fixed_pH
```

```{r echo=FALSE, Correlation_Between_pH_fixed.acidity}
# Correlation between pH fixed.acidity
cor.test(rwine$pH, rwine$fixed.acidity)
```

> It is noticible that fixed.acidity and pH have a negative correlation. When fixed.acidity increase ph levels decrease. 

```{r echo=FALSE, warning=FALSE, message=FALSE, Scatterplots_fixed.acidity_QC_Cor}
# Scatterplots between fixed.acidity and quality corelation
fixed_qc <- ggplot(aes(quality, fixed.acidity), data = rwine)+ geom_jitter() +
  xlim(3,quantile(rwine$quality, 0.99)) + 
  ylim(6,quantile(rwine$fixed.acidity, 0.99)) +
  stat_smooth(method='lm')

fixed_qc
```

```{r echo=FALSE, Correlation_Between_QC_fixed.acidity}
# Correlation between QC and fixed.acidity
cor.test(rwine$quality, rwine$fixed.acidity)
```

> There is positive correlation between fixed.acidity and quality. As the fixed.acidity increase quality gets better.

```{r echo=FALSE, warning=FALSE, message=FALSE, Scatterplots_Alcohol_QC_Cor}
# Scatterplots between Alcohol and QC Correlation
alcohol_qc <- ggplot(aes(quality, alcohol), data = rwine)+ geom_jitter()+
  xlim(3,quantile(rwine$quality, 0.99))+ 
  ylim(6,quantile(rwine$alcohol, 0.99))+
  stat_smooth(method='lm')

alcohol_qc
```

```{r echo=FALSE, warning=FALSE, message=FALSE, Correlation_Between_QC_Alcohol}
# Correlation between QCand Alcohol
cor.test(rwine$quality, rwine$alcohol)
```

> It is clear that higher the alcohol content the better the quality gets.


```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=7, fig.width=9, ggpairs}
# ggpais
rwine.subset <- rwine[,2:12]
ggpairs(rwine.subset, wrap = c(shape = I('.'), outlier.shape = I('.')))
```

> The highest positive correlatoin is between [fixed.acidity, density] and [fixed.acidity, citric.acid]

```{r echo=FALSE, Highest_Correlation}
# plot correlation between fixed.acidity and density
fixed_den <- ggplot(aes(x = fixed.acidity, y = density), data = rwine) + 
  geom_jitter(alpha = 1/5) +
  geom_smooth(method = "lm", color = "blue")

# plot correlation between fixed.acidity and citric.acid
fixed_citric <- ggplot(aes(x = fixed.acidity, y = citric.acid), data = rwine) + 
  geom_jitter(alpha = 1/5) +
  geom_smooth(method = "lm", color = "blue")

# display plot in one fig
grid.arrange(fixed_den, fixed_citric, ncol = 2)
```

> As ggpais showed us, there is a positive correlation between [fixed.acidity, density] and [fixed.acidity, citric.acid]

```{r echo=FALSE, Multiple_Correlations}

# code below will print correlation for some features
print('pH and Alcohol Correlation')
cor.test(rwine$quality, rwine$pH)

print('Density and Alcohol Correlation')
cor.test(rwine$quality, rwine$density)

print('fixed.acidity and Alcohol Correlation')
cor.test(rwine$quality, rwine$fixed.acidity)

print('fixed.acidity and density')
cor.test(rwine$fixed.acidity, rwine$density)

print('fixed.acidity and citric.acid')
cor.test(rwine$fixed.acidity, rwine$citric.acid)
```

> All plots point at those wines that fall within the categories of 5 and 6 are best. Even tho these feature are not those that consumer are looking for they also influence on the quality of the wine.

# Bivariate Analysis

## Observations regarding variables from plots: 

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

> Positive correlation [quality, alcohol], [quality, fixed.acidity], [alcohol, fixed.acidity], [fixed.acidity, citric.acid]. Consumers prefer wine with high alcohol content. 

> Negative correlation [pH, fixed.acidity], [Density, Alcohol], 

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

> ggpairs matrix, show us that there is a high positive correlation between quality and fixed.acidity, as well as between fixed.acidity and citric.acid.

### What was the strongest relationship you found?

> Alcohol and Quality.

# Multivariate Plots Section

```{r echo=FALSE, warning=FALSE, message=FALSE, Multivariate_Scatter_Plots}
# Multivariate_Scatter_Plots log(fixed.acidity) and alcohol with quality factor for the colour
fixed_log <- ggplot(aes(x = log10(fixed.acidity), y = alcohol, colour = quality.factor), 
       data = rwine) + 
  geom_point() +
  scale_color_brewer(type = 'div', palette="RdYlBu", direction = -1, name = "Quality") +
  scale_x_continuous(lim=c(quantile(log10(rwine$fixed.acidity), 0.01),
                           quantile(log10(rwine$fixed.acidity), 0.99)))+
  scale_y_continuous(lim=c(quantile(rwine$alcohol, 0.01),
                           quantile(rwine$alcohol, 0.99))) +
  ggtitle('Fixed Acidity (log10) vs Alcohol with Quality Factor')

# display plot
fixed_log

```

> The plot shows us that those wine with the highest rating fall in where the fixed.acidity and alcohol level are high. Most green dots and violet dots are found bottom left corner. 


```{r echo=FALSE, Multivariate_Scatter_Plots_1}
# # Multivariate_Scatter_Plots quality and alcohol with fixed.acidity for the colour
ggplot(aes(alcohol, quality, color=fixed.acidity), data = rwine) + 
  geom_point(alpha = 0.8, size = 2, position = 'jitter')+ 
  ggtitle('Quality vs Alcohol with Fixed Acidity content')  +
  scale_color_gradient2(midpoint = median(rwine$fixed.acidity))

```

> Wines that have higher levels of fixed.acidity have better quality than those with low fixed.acidity. 

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

> Fixed.acidity, volatile.acidity, citric.acidity, and alcohol can be used to predict the quality of the wines. Due the positive correlation I think the best thin k to do is to create a linear model to properly prdeict the wine quality.

### Were there any interesting or surprising interactions between features?

> It was kind of surprising that wine quality increase with the acidicy and alcohol levels.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

> R-square and P-value in the linear regression will help determine quality of the wine

## Regression Model

```{r echo=FALSE, Plot_One_1}
m1 <- lm(quality ~ alcohol*volatile.acidity*citric.acid*fixed.acidity, data = rwine)
m2 <- update(m1, ~ . + density*fixed.acidity)
m3 <- update(m2, ~ . + sulphates)
m4 <- update(m3, ~ . + pH*total.sulfur.dioxide*free.sulfur.dioxide)

mtable(m1, m2, m3, m4)
```


------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
# bloxplot to display relationshi between important features
grid.arrange(qc.cat_alcohol, qc.cat_sulphates, qc.cat_density, qc.cat_citric, qc.cat_fixed, ncol = 3)

```

### Description One

> Red Wine quality increase when it have high levels of alcohol, density, citric.acid, fixed.acidity. Volatile.acidity actual decrease wine quality.

### Plot Two
```{r echo=FALSE, warning=FALSE,message=FALSE, Plot_Two}
# scatterplot for quality and sulphates correaltion
sc_sulphates_qc

```

### Description Two

> There is a positive correlation between sulphates and quality. When the sulphates values increase the wine quality increases too.

### Plot Three
```{r echo=FALSE, Plot_Three}
# Quality vs Alcohol with Fixed Acidity content
ggplot(aes(alcohol, quality, color=fixed.acidity), data = rwine) + 
  geom_point(alpha = 0.8, size = 2, position = 'jitter')+ 
  ggtitle('Quality vs Alcohol with Fixed Acidity content')  +
  scale_color_gradient2(midpoint = median(rwine$fixed.acidity))
```

### Description Three

> The plot above show us that wines with cat 5 and cat 6 high values of fixed acidity.

------

# Reflection

The Red wine data set contains information about 1,599 red wines with 11 features. The analysis started by analysing individuals features in the red wine data set. The analysis revealed that alcohol, sulphates, and fixed.acidity are key features for determining wine's quality. High alcohol level and low concentration of volatile acidity are a good combination for red wine. Most frequent quality levels are cat5 and cat6. May be beauce of price too?. Exmining the correlations guided me to understand certains relations between features such as fixed.acidy and alcohol.

I always thouht that alcohol content was a key feature when measuring the quality of the wines. After the analysis, sulphates and acidity are also playing an important role in determining wine's quality. 

Some limitations, the dataset are only variants of the Portuguese "Vinho Verde" wine. There is no data about grape types, wine brand, wine selling price, etc.
